Using Occupancy Theory and Hierarchical Models to Estimate Otter Population in Southeastern Minnesota \\ {\large Senior Comprehensive Exercise: Carleton College}

### 2010-03-01 Paper.tex
Kiva's edits.

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage[dvipsnames, table]{xcolor}
\usepackage[all]{xy}
\usepackage{hyperref}

@@

% This should be a lot prettier.
\title{Using Occupancy Theory and Hierarchical Models to Estimate Otter
Population in Southeastern Minnesota \\
{\large Senior Comps Project}}
\author{Chrisna Aing, Sarah Halls, Kiva Oken}
\maketitle

\section{Questions of investigation}
The North American river otter (\textit{Lontra canadensis}) is endemic to all of
Minnesota; however, populations declined precipitously in the 19th and 20th
centuries due to trapping and habitat degradation. In recent years their
populations have significantly rebounded, especially in southeastern Minnesota
where hunters would like to be able to trap otters again. In order for that to
be possible, governmental organizations must be able to monitor the health of
the population.
In addition, otter presence can indicate the quality of the aquatic habitat.
Due to these reasons, there is interest in monitoring changes in otter
population
over time. Unfortunately, otters are not easy to observe. Thus, this study
investigated the potential of using wintertime aerial surveys of otter tracks on
rivers to make inferences about the population of otters and how it changes over
time. We used concepts from occupancy theory to develop models to estimate
the occupancy rate in order to get a sense of otter population.

    \subsection{Data collection}
    During the winters of 2003 and 2004 people flew over the Whitewater, Zumbro,
    and Mississippi Rivers. When they saw what looked like an otter track, they
    pushed a button recording the GPS coordinates. They pushed the button every
    five seconds until the track ended. Flights occurred one, two, or three days
    after a snow event. In some cases, multiple observers flew over the same
    river on the same day. Surveys were conducted during multiple snow events
    throughout both winters. The rivers were then divided into 402 and 804 meter
    plots and the presence (1) or absence (0) of a waypoint in a plot on a given
    flight was determined. The Mississippi River was also divided into 1608
    meter plots.  For each river and plot size, the data set with every other
    plot removed was considered.  We refer to this as ``alternating plots''.

@@

    In addition, there is concern that the probability of an otter laying down a
    track in one plot is not independent of the probability that an otter lays
    down a track in neighboring plots, violating a principle assumption of
    occupancy theory. That is, a single track can extend across multiple plots.
    This leads to dependence among the plots. It was important to quantify this
    dependence and find an optimum way to handle it.

@@

    \subsection{False positive rates}
    The lack of consistency among the flights is strong evidence that the data
    contain false positives where observers think they see a track when it does
    not exist. Not accounting for such a process can lead to major biases; as
    the number of flights increases, the occupancy rate estimate will approach
    1 \cite{Royle2006}. Unlike most occupancy models where observers assume
    that detection of the species in question guarantees its presence \cite
    {MacKenzie2006}, we introduced a false detection process whereby an observer
    can mark down the existence of a track when in reality there is none. This
    false detection process can occur as long as there is no track, whether or
    not the plot is occupied.

@@

\section{Spatial correlation of the data}
One approach to dealing with the correlation in the probability of
observing a track in neighboring plots is to use a conditional autoregressive
(CAR) prior distribution for $\theta$, the probability that an otter lays down a
track in a plot on a given day.

@@

    \subsection{Our CAR model}
    \label{car model}
    For our model, both neighbors are weighted equally. Since $\tau_s$ for any
    plot $s$ is unknown, we followed convention and used the same value of
    $\tau$ for all plots and placed a prior distribution on it. We chose a
    $\gamma(0.5,0.0005)$ prior distribution for $(\tau^2)^{-1}$, a common choice
    \cite{Thomas2004}. In our model we define $\text{logit}(\theta_s)=\alpha_0+
    \alpha_{1_s}$ where $\alpha_0$ has an improper uniform distribution and
    $\alpha_1$ has a Gaussian CAR distribution. We again used a uniform prior
    distribution on $\psi$ and the same beta prior distributions on $p$ and $E$.
    In this case, we limited the range of $p$ to the interval $(0.6,1)$ and the
    range of $E$ to $(0,0.3)$. We increased the burn-in time to 12,000
    iterations. With this model, instead of a single estimate for $\theta$ there
    are now $S$ estimates for $\theta$ where $S$ is the total number of plots.
    While ideally we would estimate an overall value for $\theta$, it seems more
    important for the purposes of this investigation to estimate $\psi$.

    \subsection{Simulating spatially correlated data}
    \label{spatial simulation}
    Because we modeled spatial correlation in $\theta$, we began by simulating
    spatial correlation in the track-laying process. To do this, we added an
    additional variable to the simulated data, $\alpha$. Under the new
    simulation method, $\theta$ is the probability of an otter laying down a
    track in plot $i$ given plot $i$ is occupied and plot $i-1$ does
    \textit{not} have a track. If plot $i$ is occupied and plot $i-1$ does have
    a track, then the probability of an otter laying down a track in plot $i$ is
    equal to $\theta+\alpha$. This means if there is a track in one plot, there
    is more likely to be a track in the next plot. However, when we ran data
    simulated using this new method on the model described in Section \ref{car
    model}, the model almost never converged. We tested independence of the data
    using the $\hat{S}$ statistic we developed (Section \ref{shat}) and found
    that the simulated data displayed much more independence than the actual
    data. We would like our simulated data to have the same level of
    independence
    as our actual data.  We tried modeling spatial correlation in the occupancy
    process, $\psi$, instead of in the track-laying process, but the simulated
    data were again more independent than the real data.

    To increase the correlation between plots we simulated data
    correlated in both the track-laying and occupancy processes. Under this 
    simulation method we
    again had the variables $\theta$ and $\alpha$, but we also had analogous
    variables $ \psi$ and $\beta$. In this case, $\psi$ is the probability an
    otter occupies plot $i$ given that plot $i-1$ is not occupied and $\psi+
    \beta$ is the probability that an otter occupies plot $i$ given that plot
    $i-1$ is occupied. With this increased spatial correlation we were able to
    more closely match the $\hat{S}$ Z-scores found in the actual data. In
    addition, the CAR model converged more often with data simulated using this
    method, but it still did not converge consistently. When it did converge,
    estimates were comparable to those found using the standard model on the
    same simulated data.

@@

\end{document}

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage[dvipsnames, table]{xcolor}
\usepackage[all]{xy}
\usepackage{hyperref}
\usepackage{moreverb}

@@

% This should be a lot prettier.
\title{Using Occupancy Theory and Hierarchical Models to Estimate Otter
Population in Southeastern Minnesota \\
{\large Senior Comprehensive Exercise: Carleton College}}
\author{Chrisna Aing, Sarah Halls, Kiva Oken}
\maketitle

\section{Questions of investigation}
The North American river otter (\textit{Lontra canadensis}) is endemic to all of
Minnesota; however, populations declined precipitously in the 19th and 20th
centuries due to trapping and habitat degradation. In recent years their
populations have significantly rebounded, especially in southeastern Minnesota
where hunters would like to be able to trap otters again. In order for that to
be possible, governmental organizations must be able to monitor the health of
the population.
In addition, otter presence can indicate the quality of the aquatic habitat.
Due to these reasons, there is interest in monitoring changes in otter
population
over time. Unfortunately, otters are not easy to observe. Thus, this study
investigated the potential of using wintertime aerial surveys of otter tracks on
rivers to make inferences about the population of otters and how it changes over
time. We used concepts from occupancy theory to develop models to estimate
the occupancy rate for the otter population.

    \subsection{Data collection}
    During the winters of 2003 and 2004 The Minnesota Department of Natural 
    Resources (MDNR) sponsored flights over the Whitewater, Zumbro,
    and Mississippi Rivers to survey otters. When observers saw what looked like 
    an otter track, they
    pushed a button recording on a handheld GPS unit to record the GPS 
    coordinates (waypoint). They pushed the button every
    five seconds until the track ended. Flights occurred one, two, or three days
    after a snow event. In some cases, multiple observers flew over the same
    river on the same day. Surveys were conducted during multiple snow events
    throughout both winters. The rivers were then divided into 402 and 804 meter
    plots and the presence (1) or absence (0) of a waypoint in a plot on a given
    flight was determined. The Mississippi River was also divided into 1608
    meter plots.  In addition, for each river and plot size, a data set with 
    every other
    plot removed was considered.  We refer to these as ``alternating plots''.

@@

    In addition, there is concern that the probability of an otter laying down a
    track in one plot is not independent of the probability that an otter lays
    down a track in neighboring plots. That is, a single track can extend across 
    multiple plots. This leads to dependence among the plots, violating a 
    principle assumption of many
    occupancy models. We tried to quantify this
    dependence and find a way to handle it.

@@

    \subsection{False positive rates}
    The lack of consistency among the flights is strong evidence that the data
    contain false positives where observers think they see a track when it does
    not exist. Not accounting for such a process can lead to major biases; as
    the number of flights increases, the occupancy rate estimate will approach
    1 \cite{Royle2006}. We thus introduced a false detection process whereby an 
    observer
    can mark down the existence of a track when in reality there is none. This
    false detection process can occur as long as there is no track, whether or
    not the plot is occupied.

@@

\section{Spatial correlation of the data}
One approach to dealing with the correlation in the
observed tracks in neighboring plots is to use a conditional autoregressive
(CAR) prior distribution for $\theta$, the probability that an otter lays down a
track in a plot on a given day.

@@

    \subsection{Our CAR model}
    \label{car model}
    For our model, both neighbors are weighted equally. Since $\tau_s$ for any
    plot $s$ is unknown, we followed convention and used the same value of
    $\tau$ for all plots and placed a prior distribution on it. We chose a
    $\gamma(0.5,0.0005)$ prior distribution for $(\tau^2)^{-1}$, a common choice
    \cite{Thomas2004}. In our model we define $\text{logit}(\theta_s)=\alpha_0+
    \alpha_{1_s}$ where $\alpha_0$ has an improper uniform distribution and
    $\alpha_1$ has a Gaussian CAR distribution. We again used a uniform prior
    distribution on $\psi$ and the same beta prior distributions on $p$ and $E$.
    In this case, we limited the range of $p$ to the interval $(0.6,1)$ and the
    range of $E$ to $(0,0.3)$. We increased the burn-in time to 12,000
    iterations. With this model, instead of a single estimate for $\theta$ there
    are now $n$ estimates for $\theta$ where $n$ is the total number of plots.
    While ideally we would estimate an overall value for $\theta$, it seems more
    important for the purposes of this investigation to estimate $\psi$.

    \subsection{Simulating spatially correlated data}
    \label{spatial simulation}
    We began by simulating
    spatial correlation in the track-laying process. To do this, we added an
    additional variable, $\alpha$, to the simulated data. Under the new
    simulation method, $\theta$ is the probability of an otter laying down a
    track in plot $i$ given plot $i$ is occupied and plot $i-1$ does
    \textit{not} have a track. If plot $i$ is occupied and plot $i-1$ does have
    a track, then the probability of an otter laying down a track in plot $i$ is
    equal to $\theta+\alpha$. This means if there is a track in one plot, there
    is more likely to be a track in the next plot. However, when we ran data
    simulated using this new method on the model described in Section \ref{car
    model}, the model almost never converged. We tested independence of the data
    using the $S$ statistic we developed (Section \ref{shat}) and found
    that the simulated data displayed much less dependence than the actual
    data. We would like our simulated data to have the same level of dependence
    as our actual data.  We tried modeling spatial correlation in the occupancy
    process, $\psi$, instead of in the track-laying process, but the simulated
    data were again more independent than the real data.

    To increase the correlation between plots we simulated the correlation
    in both the track-laying and occupancy processes. Under this 
    simulation method we
    again had the variables $\theta$ and $\alpha$, but we also had analogous
    variables $ \psi$ and $\beta$. In this case, $\psi$ is the probability an
    otter occupies plot $i$ given that plot $i-1$ is not occupied and $\psi+
    \beta$ is the probability that an otter occupies plot $i$ given that plot
    $i-1$ is occupied. With this increased spatial correlation we were able to
    more closely match the $S$-related Z-scores found in the actual data. In
    addition, the CAR model converged more often with data simulated using this
    method, but it still did not converge consistently. When it did converge,
    estimates were comparable to those found using the standard model on the
    same simulated data.

@@

\section{Standard hierarchical model}
\scriptsize
\begin{verbatimtab}
otterModel <- function() {
    # Initialization of variables to estimate.
    psi ~ dunif(0,1)
    theta ~ dunif(0,1)
    a ~ dgamma(1,.1)
    b ~ dgamma(1,.1)
    c ~ dgamma(1,.1)
    d ~ dgamma(1,.1)
    for (observer in 1:observers) {
        #p[observer] ~ dbeta(a,b)
        p[observer] ~ dbeta(a,b) %_% I(.5, 1)
        E[observer] ~ dbeta(c,d) %_% I(0, .5)
    }

    # Execution of all site- and track-related work.
    for (site in 1:sites) {
        z[site] ~ dbern(psi)
        lambda[site] <- theta * z[site]
        for (event in 1:snowEvents) {
            for (day in 1:daysAfterTotal) {
                track[site, event, day] ~ dbern(lambda[site])
                trackSum[site, event, day] <- sum(track[site, event, 1:day])
            }
        }
    }

    # Super difficult stuff.  (Not really.)
    for (i in 1:numDataPts) {
        # equals(track.count[...], 0) == 1 when there is no track to observe.
        mu[i] <- 
            p[obsV[i]] * (1 - equals(trackSum[siteV[i], eventV[i], dayV[i]], 0))
        # probability of false detection, given track and occupancy state
        nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
        prob[i] <- mu[i] + nu[i]
        #prob[i] <- mu[i]
        y[i] ~ dbern(prob[i])
    }
}

writeModel(otterModel, "irregularHierarchicalModel.txt")
modelCheck("irregularHierarchicalModel.txt")
bugsData(c("y", "siteV", "eventV", "dayV", "obsV", "sites", "observers",
           "numDataPts", "snowEvents", "daysAfterTotal"), 
         "irregularOtterData.txt")
modelData("irregularOtterData.txt")
modelCompile(numChains=2)
modelGenInits()

modelUpdate(10000)
samplesSet(c("a", "b", "c", "d", "p[]", "E[]", "psi", "theta"))
#samplesSet(c("a", "b", "p[]", "psi", "theta"))
modelUpdate(10000)
\end{verbatimtab}
%}

\section{CAR model}
\scriptsize
\begin{verbatimtab}
otterModel <- function() {
    for (site in 1:sites) {
        logit(theta[site]) <- alpha1[site] + alpha0
    }
    for (i in 1:sumNumNeigh) {
        weights[i] <- 1
    }
    alpha1[1:sites] ~ car.normal(adj[], weights[], num[], tau)
    alpha0 ~ dflat()
    tau ~ dgamma(.5,.0005)
    psi ~ dunif(0,1) 
    a ~ dgamma(1,.1)
    b ~ dgamma(1,.1)
    #c ~ dgamma(1,.1)
    #d ~ dgamma(1,.1)
    for (observer in 1:observers) {
        p[observer] ~ dbeta (a,b)
	  #p[observer] ~ dbeta(a,b) %_% I(.6, 1)
        #E[observer] ~ dbeta(c,d) %_% I(0, .3)
    }

    # Execution of all site- and track-related work.
    for (site in 1:sites) {
        z[site] ~ dbern(psi)
        lambda[site] <- theta[site] * z[site]
        for (event in 1:snowEvents) {
            for (day in 1:daysAfterTotal) {
                track[site, event, day] ~ dbern(lambda[site])
                trackSum[site, event, day] <- sum(track[site, event, 1:day])
            }
        }
    }

    # Super difficult stuff.  (Not really.)
    for (i in 1:numDataPts) {
        # equals(track.count[...], 0) == 1 when there is no track to observe.
        mu[i] <- 
            p[obsV[i]] * (1 - equals(trackSum[siteV[i], eventV[i], dayV[i]], 0))
        # probability of false detection, given track and occupancy state
        #nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
        #prob[i] <- mu[i] + nu[i]
        prob[i] <- mu[i]
	  y[i] ~ dbern(prob[i])
    }
}

writeModel(otterModel, "irregularHierarchicalModel.txt")
modelCheck("irregularHierarchicalModel.txt")
bugsData(c("y", "siteV", "eventV", "dayV", "obsV", "sites", "observers",
           "numDataPts", "snowEvents", "daysAfterTotal", "num", "adj",
           "sumNumNeigh"), 
         "irregularOtterData.txt")
modelData("irregularOtterData.txt")
modelCompile(numChains=2)

init.val <- function(){
	list(
		alpha1=runif(sites,-100,100),
		alpha0=runif(1,-100,100),
		tau=rgamma(1,.5,.0005),
		psi=runif(1),
		a=rgamma(1,1,.1),
		b=rgamma(1,1,.1)
		#,c=rgamma(1,1,.1),
		#d=rgamma(1,1,.1)
	) 
}
bugsInits(init.val,numChains=2,fileName=c('init1.txt','init2.txt'))
modelInits(c('init1.txt','init2.txt'))

modelGenInits()

modelUpdate(12000)
#samplesSet(c("a", "b", "c", "d", "p[]", "E[]", "psi"))
samplesSet(c("a", "b", "p[]", "psi"))
modelUpdate(10000)
\end{verbatimtab}

\section{Simulated data, independent plots}
\scriptsize
\begin{verbatimtab}
# Setting up the environment and results matrix.
sites <- 100
snowEvents <- 2
daysAfterTotal <- 2
observers <- 3

eventsToDaysBegin <- c(1, 3, 5, 7, 9, 11)
eventsToDaysEnd   <- c(2, 4, 6, 8, 10, 12)
curDaySinceSnow <- c(1, 2, 1, 2, 1, 2, 1, 2, 1, 2, 1, 2)
daysToObsBegin <- c(1, 4, 7, 10, 13, 16, 19, 22, 25, 28, 31, 34)
daysToObsEnd   <- c(3, 6, 9, 12, 15, 18, 21, 24, 27, 30, 33, 36)
curObserver <- c(1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3,
                 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3)

# Setting up the parameters.
psi <- .7
beta <- .4
condlOcc <- psi + beta
condlUnocc <- psi

theta <- .5
alpha <- .3
condlTrack <- theta + alpha
condlNoTrack <- theta
p <- c(.70, .80, .90)
E <- c(.05, .02, .03)

# Retaining necessary information.
occupancy <- NULL

# Vectors; employing a nested index strategy to deal with WinBUGS and its lack
# of if-then statements.
y <- NULL
siteV <- NULL
eventV <- NULL
dayV <- NULL
obsV <- NULL
numDataPts <- 0

# Simulating the data!
for (site in 1:sites) {
    # TRUE == 1, FALSE == 0.
	if (site == 1) {
		occupancy[site] <- 1 * (runif(1) < (condlOcc + condlUnocc/2))
	} else if (occupancy[i-1] == 1) {
		occupancy[site] <- 1 * (runif(1) < condlOcc)
	} else
		occupancy[site] <- 1 * (runif(1) < condlUnocc)
	}
    for (event in 1:snowEvents) {
        otterTrackLaid <- 0
        currentDay <- 0
        # Haha, this is awful.
        for (day in 1:curDaySinceSnow[eventsToDaysBegin[event]]) {
            if (day != 1) {
                currentDay <- currentDay + 1
                otterTrackLaid <- max(otterTrackLaid,
                    occupancy[site] * (runif(1) < theta))
            }
        }
        for (day in eventsToDaysBegin[event]:eventsToDaysEnd[event]) {
            currentDay <- currentDay + 1
            otterTrackLaid <- max(otterTrackLaid,
                occupancy[site] * (runif(1) < theta))
            for (colIndex in daysToObsBegin[day]:daysToObsEnd[day]) {
                numDataPts <- numDataPts + 1
                sawOtterTrack <- otterTrackLaid * 
                    (runif(1) < p[curObserver[colIndex]])
                sawOtherTrack <- (1 - otterTrackLaid) * 
                    (runif(1) < E[curObserver[colIndex]])
                y[numDataPts] <- sawOtterTrack + sawOtherTrack
                siteV[numDataPts] <- site
                eventV[numDataPts] <- event
                dayV[numDataPts] <- currentDay
                obsV[numDataPts] <- curObserver[colIndex]
            }
        }
    }
}
\end{verbatimtab}

\section{Simulated regular data, dependent in $\psi$, $\theta$}
\scriptsize
\begin{verbatimtab}
#Define structure
sites <- 60
events <- 5
snowdays <- 3
obs <- 3
I1 <- rep(c(1,0,0), events*snowdays)
I2 <- rep(c(0,1,0), events*snowdays)

#Define parameters
psi <- .4
beta <- .4
condl.occ <- psi + beta
condl.unocc <- psi
theta <-.3
alpha <-.5
condl.track <- theta + alpha
condl.notrack <- theta
p <- c(.7,.7,.7)
E <- c(.05,.05,.05)

#Initialize variables
occupancy <- NULL
tracks <- matrix(nrow=sites, ncol=events*snowdays)
y <- matrix(nrow=sites, ncol=events*snowdays*obs)

for(i in 1:sites) {
	#Determine occupancy of site. 1*(runif(1)<psi) is 1 around psi*100 percent of the time.
	#Use conditional rates depending on occupancy of site i-1
	if (i==1) 
		{occupancy[i] <- 1*(runif(1)<(condl.occ+condl.unocc)/2)}	
	else if (occupancy[i-1]==1) 
		{occupancy[i] <- 1*(runif(1)<condl.occ)}
	else 
		{occupancy[i] <- 1*(runif(1)<condl.unocc)}
	for(j in 1:events) {
		is.track <- 0
		for(k in 1:snowdays) {
			index <- k + snowdays*(j-1)
			#Decide if a track is laid down on day k of event j, given the occupancy state
			if(i==1) {
				tracks[i,index] <- (runif(1)<(condl.track+condl.notrack)/2)*occupancy[i]
			}else if(tracks[i-1,index]==1) 
				{tracks[i,index] <- (runif(1)<condl.track)*occupancy[i]
			}else 
				{tracks[i,index] <- (runif(1)<condl.notrack)*occupancy[i]}

			#Determine if any tracks have been laid down for event j.
			is.track <- max(is.track, tracks[i,index])
			for(m in 1:obs) {
				index <- m + obs*(k-1) + snowdays*obs*(j-1)
				#Determine if the observer saw a track.
				y[i,index] <- (runif(1)<p[m])*is.track + (runif(1)<E[m])*(1-is.track)
			}
		}
	}
}
\end{verbatimtab}
Note: This needs some more processing before it will run on either model.
\end{document}
